(Re-)bridging the anatomy ontologies with SSSOM

### 2024-05-28 presentation.tex
Rework "current system" and "conclusion".

\begin{frame}
  \frametitle{Current solution: cross-references}

@@

xref: FBbt:00003154
  \end{block}
\end{frame}

@@

  \begin{columns}
    \begin{column}{.3\textwidth}
      \begin{tikzpicture}
        \node [File] (uberon) at (0,0)     {uberon.obo};
        \node [File] (script) at (0,-1)    {make-bridge-ontology-from-xrefs.pl};
        \node [File] (bridge) at (0,-2)    {uberon-bridge-to-fbbt.obo};
        \node [File]          at (0, -2.5) {uberon-bridge-to-wbbt.obo};
        \node [File]          at (0, -3)   {uberon-bridge-to-ma.obo};
        \node [File]          at (0, -3.5) {uberon-bridge-to-....obo}; 
        \draw[->] (uberon) -- (script) (script) -- (bridge);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.68\textwidth}
      \begin{block}{Bridge file}\ttfamily
[Term]

@@

property\_value: IAO:0000589 "heart (drosophila)" xsd:string

@@

intersection\_of: part\_of NCBITaxon:7227
      \end{block}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}
  \frametitle{Limitations of the current system}

  Limitations of cross-references to represent mappings
  \begin{itemize}
    \item Overloading of the \emph{oboInOwl:hasDbXref} annotation
    \item No metadata
    \item Mappings are embedded within the ontology and must be extracted
  \end{itemize}

  \bigskip
  Limitations of the bridging pipeline
  \begin{itemize}
    \item Highly Uberon-specific -- cannot be reused outside of Uberon
    \item Dependent on the OBO format
  \end{itemize}
\end{frame}

@@

  \begin{enumerate}
    \item Clear separation between cross-ontology \emph{mappings} (COM) and cross-ontology \emph{bridges} (COB)
      \begin{itemize}
        \item A COB is merely the OWL materialisation of a COM
      \end{itemize}
    \item Represent COM using a standard format intended for that purpose
      \begin{itemize}
        \item allowing for mapping-specific metadata
        \item allowing for meaningful mapping predicates
      \end{itemize}
    \item Make it easy to derive COB from COM
      \begin{itemize}
        \item without requiring Uberon-specific tooling
      \end{itemize}
  \end{enumerate}
\end{frame}

@@

\begin{frame}
  \frametitle{Integrating the species-specific ontologies}

  \begin{center}
    \begin{tikzpicture}[node distance=.5cm]
      \node [Entity] (uberon) {Uberon};
      \node [Entity,right=of uberon] (com) {Mapping sets};
      \node [Entity,right=of com] (ssao) {Taxon-specific ontologies};
      \node [Process,below=of com] (sssomt) {SSSOM/T-OWL};
      \node [Entity,below=of sssomt] (cob) {Bridge files};
      \node [Process,below=of cob] (merge) {OWL Merge};
      \node [Entity,below=of merge] (cm) {Multi-species ontology};
      \draw [->] (com) -- (sssomt) (sssomt) -- (cob);
      \draw [->] (cob) -- (merge) (merge) -- (cm);
      \draw [->] (uberon.south) |- (merge.west);
      \draw [->] (ssao.south) |- (merge.east);
    \end{tikzpicture}
  \end{center}
\end{frame}

@@

\begin{frame}
  \frametitle{Outcome}

>>>>

\begin{frame}
  \frametitle{Maintaining mappings as cross-references}

@@

xref: FBbt:00003154
  \end{block}

  Limitations:
  \begin{itemize}
    \item Overloading of the \emph{oboInOwl:hasDbXref} annotation
    \item No metadata
    \item Mappings are embedded within the ontology and must be extracted
  \end{itemize}
\end{frame}

@@

  \begin{columns}
    \begin{column}{.3\textwidth}
      \begin{tikzpicture}
        \node [File] (uberon) at (0,0)     {uberon.obo};
        \node [File] (script) at (0,-1)    {make-bridge-ontology-from-xrefs.pl};
        \node [File] (bridge) at (0,-2)    {uberon-bridge-to-fbbt.obo};
        \node [File]          at (0, -2.5) {uberon-bridge-to-wbbt.obo};
        \node [File]          at (0, -3)   {uberon-bridge-to-ma.obo};
        \node [File]          at (0, -3.5) {uberon-bridge-to-....obo}; 
        \draw[->] (uberon) -- (script) (script) -- (bridge);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.68\textwidth}
      \begin{block}{uberon-bridge-to-fbbt.obo}\ttfamily
[Term]

@@

@@

intersection\_of: part\_of NCBITaxon:7227
      \end{block}

      Can be used to integrate FBbt into Uberon
    \end{column}
  \end{columns}

  \medskip
  Limitations:
  \begin{itemize}
    \item Highly Uberon-specific code -- cannot be reused outside of Uberon
    \item Dependent on the OBO format
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Integration pipeline}

  \begin{columns}
    \begin{column}{.44\textwidth}
      \begin{tikzpicture}[node distance=.5cm]
        \node [Entity] (uberon) {Uberon};
        \node [Process,below right=of uberon] (script) {Custom script};
        \node [Entity,above right=of script,align=left] (ssao) {Taxon-specific\\
                                                                ontologies};
        \node [Entity,below=of script] (cob) {Bridge files};
        \node [Process,below=of cob] (merge) {OWL merge};
        \node [Entity,below=of merge] (cm) {Multi-species ontology};
        \draw [->] (uberon.east) -| (script.north) (script.south) -- (cob.north);
        \draw [->] (cob) -- (merge) (merge) -- (cm);
        \draw [->] (uberon.south) |- (merge.west);
        \draw [->] (ssao.south) |- (merge.east);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.55\textwidth}
      Limitations:
      \begin{itemize}
        \item Lack of flexibility
          \begin{itemize}
            \item Cannot integrate foreign mappings
            \item Cannot generate a single bridge file
          \end{itemize}
        \item Mappings not available independently
        \item Not generalizable to other ontologies
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

@@

  \begin{enumerate}
    \item Clear separation between
      \begin{itemize}
        \item cross-ontology \emph{mappings} (COM)
        \item cross-ontology \emph{bridges} (COB)
      \end{itemize}
    \item Represent COM using a standard format intended for that purpose
      \begin{itemize}
        \item allowing for mapping-specific metadata
        \item allowing for meaningful mapping predicates
      \end{itemize}
    \item Make it easy to derive COB from COM
      \begin{itemize}
        \item without requiring Uberon-specific tooling
      \end{itemize}
  \end{enumerate}
\end{frame}

@@

\begin{frame}
  \frametitle{New SSSOM-based integration pipeline}

  \begin{center}
    \begin{tikzpicture}[node distance=.5cm]
      \node [Entity] (uberon) {Uberon};
      \node [Entity,right=of uberon] (com) {Mapping sets};
      \node [Entity,right=of com,align=left] (ssao) {Taxon-specific\\
                                                     ontologies};
      \node [Process,below=of com] (sssomt) {SSSOM/T-OWL};
      \node [Entity,below=of sssomt] (cob) {Bridge files};
      \node [Process,below=of cob] (merge) {OWL merge};
      \node [Entity,below=of merge] (cm) {Multi-species ontology};
      \draw [->] (com) -- (sssomt) (sssomt) -- (cob);
      \draw [->] (cob) -- (merge) (merge) -- (cm);
      \draw [->] (uberon.south) |- (merge.west);
      \draw [->] (ssao.south) |- (merge.east);
    \end{tikzpicture}
  \end{center}
\end{frame}

@@

\begin{frame}
  \frametitle{Benefits of the new approach}
